---
<<<<<<< Updated upstream
title: "Template" Hoe zwaar weegt welvaart?
=======
title: ""
>>>>>>> Stashed changes
author: "Studentnames and studentnumbers here" Tobias Pierik: 2898032, Meike Bliek: 2892575, Vycho de Graaf: 2892702, Nynke Latham: 2862763, Anne Mak: 2898488
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("yaml")
#install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Title Page
<<<<<<< Updated upstream
<<<<<<< HEAD

Include your names, Anne Mak, Nynke Latham, Tobias Pierik, Vycho de Graaf, Meike Bliek


<<<<<<< HEAD
=======
>>>>>>> Stashed changes
Include the tutorial group number: 1
=======
Include the tutorial group number: 1,8
>>>>>>> 261c22b67f2673b22462cb5c7f71ad44aefb7a9c

Include your tutorial lecturer's name: Fitzgerald, J.F.

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem
In welke mate welvaart overgewicht beinvloed binnen Europa.
Include the following:

-   Why is this relevant?
Dit is een relevant onderwerp omdat overgewicht zorgt voor gezondheidsproblemen. Door onderzoek te doen naar de correlatie tussen rijkdom en overgewicht, kunnen we beter begrijpen hoe inkomen de gezondheid in een land bepaald. Hierdoor kunnen we maatregelen nemen om overgewicht te voorkomen.
-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data
https://ec.europa.eu/eurostat/databrowser/view/ilc_hch10/default/table?lang=en&category=livcon.ilc.ilc_ahm.ilc_hch 

```{r}
library(readr)
overgewichtLanden <- read_csv("data/overgewichtLanden.csv")
#View(overgewichtLanden)


library(readxl)
earn_ses_annual <- read_excel("data/earn_ses_annual.xlsx")

# kopjes namen veranderen en kopjes verwijderen  
earn_ses_annual$data2014 <- earn_ses_annual$...8
earn_ses_annual$...8 <- NULL
earn_ses_annual$data2022 <- earn_ses_annual$...12
earn_ses_annual$...12 <- NULL
earn_ses_annual$data2018 <- earn_ses_annual$...10
earn_ses_annual$...10 <-NULL

# de eerste 11 rijen weghalen earn clean is die zonder de rijen
earn_clean <- earn_ses_annual[-c(1:11),]

write.csv(earn_clean, "data/clean_earn_ses.csv")

inkomen_landen_europa <- read_csv("data/clean_earn_ses.csv")
# data zien
View(inkomen_landen_europa)
inkomen_landen_europa$data2010 <- inkomen_landen_europa$...7
inkomen_landen_europa$...7 <-NULL
inkomen_landen_europa$data2006 <- inkomen_landen_europa$...5
inkomen_landen_europa$...5 <-NULL
inkomen_landen_europa$data2002 <- inkomen_landen_europa$...3
inkomen_landen_europa$...3 <-NULL
#colommen verwijderen
inkomen_landen_europa$...4 <-NULL
inkomen_landen_europa$...6 <-NULL
inkomen_landen_europa$...8 <-NULL
inkomen_landen_europa$...9 <-NULL
inkomen_landen_europa$...10 <-NULL
inkomen_landen_europa$...11 <-NULL

#rijen weghalen
inkomen_landen_europa <- inkomen_landen_europa[-c(47:49),]

#rijen weghalen
overgewichtLanden <-overgewichtLanden [-c(1:10),]
overgewichtLanden <-overgewichtLanden [-c(43:49),]

#namen overgewicht data veranderen 
overgewichtLanden$data2017 <- overgewichtLanden$...3
overgewichtLanden$...3 <-NULL
overgewichtLanden$data2022 <- overgewichtLanden$...5
overgewichtLanden$...5 <-NULL
overgewichtLanden$data2025 <- overgewichtLanden$...7
overgewichtLanden$...7 <-NULL

#colommen weghalen
overgewichtLanden$...8 <-NULL
overgewichtLanden$...6 <-NULL
overgewichtLanden$...4 <-NULL

#rij weghalen
overgewichtLanden <-overgewichtLanden [-c(1),]
inkomen_landen_europa <- inkomen_landen_europa[-c(1),]

#colommen hernoemen
overgewichtLanden$regio <- overgewichtLanden$`Data extracted on 03/06/2026 14:05:17 from [ESTAT]`
overgewichtLanden$`Data extracted on 03/06/2026 14:05:17 from [ESTAT]` <-NULL
inkomen_landen_europa$regio <- inkomen_landen_europa$`Data extracted on 03/06/2026 14:26:04 from [ESTAT]`
inkomen_landen_europa$`Data extracted on 03/06/2026 14:26:04 from [ESTAT]`<-NULL

#Rijen weghalen
inkomen_landen_europa <- inkomen_landen_europa[-c(1:8),]
overgewichtLanden <-overgewichtLanden [-c(1:4),]

#variabelen unieke namen geven
overgewichtLanden$overgewicht2017 <- overgewichtLanden$data2017
overgewichtLanden$data2017 <-NULL
overgewichtLanden$overgewicht2022 <- overgewichtLanden$data2022
overgewichtLanden$data2022<-NULL
overgewichtLanden$overgewicht2025 <- overgewichtLanden$data2025
overgewichtLanden$data2025<-NULL




inkomen_landen_europa$inkomen2002 <- inkomen_landen_europa$data2002
inkomen_landen_europa$data2002 <-NULL
inkomen_landen_europa$inkomen2006 <- inkomen_landen_europa$data2006
inkomen_landen_europa$data2006 <-NULL
inkomen_landen_europa$inkomen2010 <- inkomen_landen_europa$data2010
inkomen_landen_europa$data2010 <-NULL
inkomen_landen_europa$inkomen2014 <- inkomen_landen_europa$data2014
inkomen_landen_europa$data2014 <-NULL
inkomen_landen_europa$inkomen2018 <- inkomen_landen_europa$data2018
inkomen_landen_europa$data2018 <-NULL
inkomen_landen_europa$inkomen2022 <- inkomen_landen_europa$data2022
inkomen_landen_europa$data2022 <-NULL

#landen met NA filteren
overgewichtLanden <- overgewichtLanden %>% filter(regio != "United Kingdom")
inkomen_landen_europa <- inkomen_landen_europa %>% filter(regio != "United Kingdom")
overgewichtLanden <- overgewichtLanden %>% filter(regio != "Montenegro")
inkomen_landen_europa <- inkomen_landen_europa %>% filter(regio != "Montenegro")

# datasets samenvoegen op regio
combined <- merge(inkomen_landen_europa, overgewichtLanden, by = "regio")
require(tidyverse)
require(ggplot2)
# scatterplot
ggplot(combined, aes(x = as.numeric(inkomen2022), y = as.numeric(overgewicht2022))) +
  geom_point() +
  geom_text(
    label = combined$regio,
    nudge_x = 1.0, nudge_y = 1.0,
    check_overlap = T
  ) +
  labs(
    x = "Inkomen 2022",
    y = "Overgewicht 2022",
    title = "Inkomen vs overgewicht per Europees land"
  ) +
  theme(axis.text.x = element_text(angle = 45))
#View(combined)
library(tidyverse)
library(data.table)

filter(combined, combined$overgewicht2022 == ":")

combined <- replace(combined, combined == ":", NA)
#grafiek verbreden
png("combined.png", width = 1500, height = 600)
plot(combined$inkomen2022, combined$overgewicht2022)
dev.off()
```

```{r}
# 1. Pakketten laden
install.packages("viridis")   # alleen de eerste keer nodig
library(viridis)
library(tidyverse)            # bevat ggplot2, dplyr, tidyr

# 2. Inkomensgroep aanmaken in 'combined'
combined <- combined %>%
  mutate(
    inkomen2022 = as.numeric(inkomen2022),
    groep = case_when(
      inkomen2022 < 20000                         ~ "Arm",
      inkomen2022 >= 20000 & inkomen2022 < 60000  ~ "Midden",
      inkomen2022 >= 60000                        ~ "Rijk"
    )
  )

# 3. Data van breed naar lang omzetten
boxplot_long <- combined %>%
  select(regio, groep, overgewicht2017, overgewicht2022, overgewicht2025) %>%
  pivot_longer(
    cols = c(overgewicht2017, overgewicht2022, overgewicht2025),
    names_to = "jaar",
    values_to = "overgewicht"
  ) %>%
  mutate(
    overgewicht = as.numeric(overgewicht),
    jaar = recode(jaar,
      "overgewicht2017" = "2017",
      "overgewicht2022" = "2022",
      "overgewicht2025" = "2025"
    )
  )

# 4. De boxplot
boxplot_long %>%
  filter(!is.na(groep)) %>%
  mutate(groep = factor(groep, levels = c("Arm", "Midden", "Rijk"))) %>%
  ggplot(aes(x = groep, y = overgewicht, fill = groep)) +
    geom_boxplot() +
    facet_wrap(~jaar) +
    scale_fill_viridis(discrete = TRUE, alpha = 0.6, option = "A") +
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 11)
    ) +
    ggtitle("BMI per inkomensgroep, opgesplitst per jaar") +
    xlab("") +
    ylab("BMI")          # <- dit zet "BMI" op de y-as
```


```{r}
#combined <- merge(inkomen_landen_europa, overgewichtLanden, by = "regio")

# werken aan nieuwe grafiek
# Library
library(cartography)
library(sp)

# Upload data attached with the package.
data(nuts2006)

# Now we have a geospatial object called nuts2.spdf containing the shape of european regions. We can plot it with the plot function.
# summary(nuts2.spdf)

# We also have a dataframe with information concerning every region.
# head(nuts2.df)
# Both object have a first column "id" that makes the link between them.

# Plot Europe
plot(nuts0.spdf, border = NA, col = NA, bg = "#A6CAE0")
plot(world.spdf, col = "#E3DEBF", border = NA, add = TRUE)
plot(nuts0.spdf, col = "#D1914D", border = "grey80",  add = TRUE)

# Add circles proportional to the total population
propSymbolsLayer(spdf = nuts0.spdf, df = nuts0.df,
    var = "pop2008", symbols = "circle", col = "#920000",
    legend.pos = "right", legend.title.txt = "Total\npopulation (2008)",
    legend.style = "c")

# Add titles, legend...
layoutLayer(title = "Countries Population in Europe",
    author = "cartography", sources = "Eurostat, 2008",
    scale = NULL, south = TRUE)


library(cartography)
library(sp)
library(dplyr)

iso_tabel <- data.frame(
  regio = c(
    "Belgium","Bulgaria","Czechia","Denmark","Germany","Estonia",
    "Ireland","Greece","Spain","France","Croatia","Italy",
    "Cyprus","Latvia","Lithuania","Luxembourg","Hungary","Malta",
    "Netherlands","Austria","Poland","Portugal","Romania",
    "Slovenia","Slovakia","Finland","Sweden","Iceland",
    "Norway","Switzerland","Bosnia and Herzegovina",
    "North Macedonia","Serbia","Türkiye"
  ),
  id = c(
    "BE","BG","CZ","DK","DE","EE","IE","EL","ES","FR",
    "HR","IT","CY","LV","LT","LU","HU","MT","NL","AT",
    "PL","PT","RO","SI","SK","FI","SE","IS","NO","CH",
    "BA","MK","RS","TR"
  )
)

combined <- merge(combined, iso_tabel, by = "regio")

# Cartography data laden
data(nuts2006)

# Zorg dat de landnamen overeenkomen
kaartdata <- nuts0.spdf %>%
  left_join(combined, by = "id")

# Variabele numeriek maken
kaartdata$overgewicht2022 <- as.numeric(kaartdata$overgewicht2022)

# Kaart tekenen

kaartdata$overgewicht2022 <- kaartdata$overgewicht2022 %>% 
  as.numeric()

# hieronder komt ggplot code
plot(nuts0.spdf,
     border = "grey70",
     col = "white")

choroLayer(
  spdf = nuts0.spdf,
  df = kaartdata,
  var = "overgewicht2022",
  col = carto.pal(pal1 = "orange.pal"),
  border = "grey50",
  legend.pos = "right",
  legend.title.txt = "Overgewicht (%) 2022"
)

layoutLayer(
  title = "Overgewicht per Europees land (2022)",
  sources = "Eurostat",
  author = "Jullie namen"
)


ggplot(kaartdata) +
  geom_sf(aes(fill = overgewicht2022)) +
  theme_void()
```

```{r}

```


```{r loading_data}
library(tidyverse)
dataset <- midwest
filter(dataset, state == "IL")
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)
De eerste dataset toont de ontwikkeling van het inkomen van de bevolking in verschillende Europese landen tussen 2002 en 2022. Over het algemeen is het inkomen in de meeste landen gestegen gedurende deze periode, hoewel de groei niet overal even sterk was. Daarnaast zijn er duidelijke verschillen zichtbaar tussen landen, wat wijst op uiteenlopende economische ontwikkelingen en welvaartsniveaus binnen Europa.
De tweede dataset geeft informatie over het percentage van de bevolking met overgewicht in verschillende Europese landen in 2017, 2020 en 2025. De gegevens laten zien dat overgewicht in veel landen een veelvoorkomend gezondheidsprobleem is. In meerdere landen is een stijgende trend zichtbaar, wat erop wijst dat het aandeel mensen met overgewicht in de loop der jaren verder toeneemt. Hierdoor blijft overgewicht een belangrijk aandachtspunt voor de volksgezondheid in Europa.

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
