---
title: "Hoe zwaar weegt welvaart"
author: "'Tobias Pierik: 2898032, Meike Bliek: 2892575, Vycho de Graaf: 2892702, Nynke Latham: 2862763, Anne Mak: 2898488'"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

# Set-up your environment

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("yaml")
#install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Title Page

#introduction
In this document, we will examine income and the percentage of overweight people across Europe. We are curious to see if a correlation can be found between a country's income and the percentage of overweight people. We started by adding two datasets, then cleaned and merged them into a single combined set. Following this, we created four graphs showing both the correlation between the two variables and the distribution of income and the percentage of overweight people across Europe.


Include your tutorial lecturer's name: Fitzgerald, J.F.

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem
Social question: To what extent does income influence the percentage of overweight people in Europe?

This is a relevant topic because being overweight causes health problems (Zorginstituut Nederland, n.d.). By researching the correlation between wealth and being overweight, we can better understand how income determines health in a country. This gives us a clearer overview of whether there is a clear monetary cause in the national percentage of people who are overweight. We assume in advance that a high income enables men to eat more healthily and can thus largely prevent being overweight (De Witte, 2025).. 
-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data
Eurostat. (n.d.). Obesity rate by body mass index (BMI) (sdg_02_10) [Data set]. Geraadpleegd op 2 juni 2026, van https://ec.europa.eu/eurostat/databrowser/view/sdg_02_10/default/table?lang=en

Eurostat. (2024, 7 november). Annual full-time adjusted salary in EU grew in 2023. European Commission. https://ec.europa.eu/eurostat/en/web/products-eurostat-news/w/DDN-20241107-1



## 2.2 Provide a short summary of the dataset(s)
The first dataset shows the development of population income in various European countries between 2002 and 2022. In general, income increased in most countries during this period, although growth was not equally strong everywhere. Additionally, clear differences are visible between countries, indicating divergent economic developments and levels of prosperity within Europe.
The second dataset provides information on the percentage of the population with overweight in various European countries in 2017, 2020, and 2025. The data show that being overweight is a common health problem in many countries. An upward trend is visible in several countries, indicating that the proportion of people with overweight continues to increase over the years. As a result, being overweight remains a major public health concern in Europe.

```{r}
library(readr)
overgewichtLanden <- read_csv("data/overgewichtLanden.csv")
#View(overgewichtLanden)


library(readxl)
earn_ses_annual <- read_excel("data/earn_ses_annual.xlsx")

# kopjes namen veranderen en kopjes verwijderen  
earn_ses_annual$data2014 <- earn_ses_annual$...8
earn_ses_annual$...8 <- NULL
earn_ses_annual$data2022 <- earn_ses_annual$...12
earn_ses_annual$...12 <- NULL
earn_ses_annual$data2018 <- earn_ses_annual$...10
earn_ses_annual$...10 <-NULL

# de eerste 11 rijen weghalen earn clean is die zonder de rijen
earn_clean <- earn_ses_annual[-c(1:11),]

write.csv(earn_clean, "data/clean_earn_ses.csv")

inkomen_landen_europa <- read_csv("data/clean_earn_ses.csv")
# data zien
View(inkomen_landen_europa)
inkomen_landen_europa$data2010 <- inkomen_landen_europa$...7
inkomen_landen_europa$...7 <-NULL
inkomen_landen_europa$data2006 <- inkomen_landen_europa$...5
inkomen_landen_europa$...5 <-NULL
inkomen_landen_europa$data2002 <- inkomen_landen_europa$...3
inkomen_landen_europa$...3 <-NULL
#colommen verwijderen
inkomen_landen_europa$...4 <-NULL
inkomen_landen_europa$...6 <-NULL
inkomen_landen_europa$...8 <-NULL
inkomen_landen_europa$...9 <-NULL
inkomen_landen_europa$...10 <-NULL
inkomen_landen_europa$...11 <-NULL

#rijen weghalen
inkomen_landen_europa <- inkomen_landen_europa[-c(47:49),]

#rijen weghalen
overgewichtLanden <-overgewichtLanden [-c(1:10),]
overgewichtLanden <-overgewichtLanden [-c(43:49),]

#namen overgewicht data veranderen 
overgewichtLanden$data2017 <- overgewichtLanden$...3
overgewichtLanden$...3 <-NULL
overgewichtLanden$data2022 <- overgewichtLanden$...5
overgewichtLanden$...5 <-NULL
overgewichtLanden$data2025 <- overgewichtLanden$...7
overgewichtLanden$...7 <-NULL

#colommen weghalen
overgewichtLanden$...8 <-NULL
overgewichtLanden$...6 <-NULL
overgewichtLanden$...4 <-NULL

#rij weghalen
overgewichtLanden <-overgewichtLanden [-c(1),]
inkomen_landen_europa <- inkomen_landen_europa[-c(1),]

#colommen hernoemen
overgewichtLanden$regio <- overgewichtLanden$`Data extracted on 03/06/2026 14:05:17 from [ESTAT]`
overgewichtLanden$`Data extracted on 03/06/2026 14:05:17 from [ESTAT]` <-NULL
inkomen_landen_europa$regio <- inkomen_landen_europa$`Data extracted on 03/06/2026 14:26:04 from [ESTAT]`
inkomen_landen_europa$`Data extracted on 03/06/2026 14:26:04 from [ESTAT]`<-NULL

#Rijen weghalen
inkomen_landen_europa <- inkomen_landen_europa[-c(1:8),]
overgewichtLanden <-overgewichtLanden [-c(1:4),]

#variabelen unieke namen geven
overgewichtLanden$overgewicht2017 <- overgewichtLanden$data2017
overgewichtLanden$data2017 <-NULL
overgewichtLanden$overgewicht2022 <- overgewichtLanden$data2022
overgewichtLanden$data2022<-NULL
overgewichtLanden$overgewicht2025 <- overgewichtLanden$data2025
overgewichtLanden$data2025<-NULL




inkomen_landen_europa$inkomen2002 <- inkomen_landen_europa$data2002
inkomen_landen_europa$data2002 <-NULL
inkomen_landen_europa$inkomen2006 <- inkomen_landen_europa$data2006
inkomen_landen_europa$data2006 <-NULL
inkomen_landen_europa$inkomen2010 <- inkomen_landen_europa$data2010
inkomen_landen_europa$data2010 <-NULL
inkomen_landen_europa$inkomen2014 <- inkomen_landen_europa$data2014
inkomen_landen_europa$data2014 <-NULL
inkomen_landen_europa$inkomen2018 <- inkomen_landen_europa$data2018
inkomen_landen_europa$data2018 <-NULL
inkomen_landen_europa$inkomen2022 <- inkomen_landen_europa$data2022
inkomen_landen_europa$data2022 <-NULL

#landen met NA filteren
overgewichtLanden <- overgewichtLanden %>% filter(regio != "United Kingdom")
inkomen_landen_europa <- inkomen_landen_europa %>% filter(regio != "United Kingdom")
overgewichtLanden <- overgewichtLanden %>% filter(regio != "Montenegro")
inkomen_landen_europa <- inkomen_landen_europa %>% filter(regio != "Montenegro")

# datasets samenvoegen op regio
combined <- merge(inkomen_landen_europa, overgewichtLanden, by = "regio")

```

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.


Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1, echo=FALSE}
#gemiddelde inkomen 
library(sf)
library(dplyr)
resultaat_inkomen <- combined %>%
   st_drop_geometry() %>%
  summarise(
    INCgem2002 = mean(suppressWarnings(as.numeric(inkomen2002, ":")), na.rm = TRUE),
    INCgem2006 = mean(suppressWarnings(as.numeric(inkomen2006, ":")), na.rm = TRUE),
    INCgem2010 = mean(suppressWarnings(as.numeric(inkomen2010, ":")), na.rm = TRUE),
    INCgem2014 = mean(suppressWarnings(as.numeric(inkomen2014, ":")), na.rm = TRUE),
    INCgem2018 = mean(suppressWarnings(as.numeric(inkomen2018, ":")), na.rm = TRUE),
    INCgem2022 = mean(suppressWarnings(as.numeric(inkomen2022, ":")), na.rm = TRUE)
  )
resultaat_inkomen
```
```{r}
groei_inkomen <- resultaat_inkomen %>%
  mutate(
    groei_2002_2006 = (INCgem2006 - INCgem2002) / INCgem2002 * 100,
    groei_2006_2010 = (INCgem2010 - INCgem2006) / INCgem2006 * 100,
    groei_2010_2014 = (INCgem2014 - INCgem2010) / INCgem2010 * 100,
    groei_2014_2018 = (INCgem2018 - INCgem2014) / INCgem2014 * 100,
    groei_2018_2022 = (INCgem2022 - INCgem2018) / INCgem2018 * 100
  ) %>%
  select(starts_with("groei_"))

groei_inkomen
```

Variable 2

```{r gen_var2, echo=FALSE}
#gemiddelde overgewicht
library(sf)
library(dplyr)
resultaat_overgewicht <- combined %>%
   st_drop_geometry() %>%
  summarise(
    OVgem2017 = mean(suppressWarnings(as.numeric(overgewicht2017, ":")), na.rm = TRUE),
    OVgem2022 = mean(suppressWarnings(as.numeric(overgewicht2022, ":")), na.rm = TRUE),
    OVgem2025 = mean(suppressWarnings(as.numeric(overgewicht2025, ":")), na.rm = TRUE)
  )
resultaat_overgewicht
```
```{r echo=FALSE}
groei_overgewicht <- resultaat_overgewicht %>%
  mutate(
    groei_2017_2022 = (OVgem2022 - OVgem2017) / OVgem2017 * 100,
    groei_2022_2025 = (OVgem2025 - OVgem2022) / OVgem2022 * 100
  ) %>%
  select(starts_with("groei_"))

groei_overgewicht
```

## 3.3 Visualize temporal variation

```{r echo=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)

# inkomen: lang formaat + eindjaar van de periode als getal
groei_inkomen_lang <- groei_inkomen %>%
  pivot_longer(everything(), names_to = "periode", values_to = "groei") %>%
  mutate(
    variabele = "Income",
    jaar = as.numeric(sub(".*_", "", periode))   # pakt het jaar na de laatste _
  )

# overgewicht: idem
groei_overgewicht_lang <- groei_overgewicht %>%
  pivot_longer(everything(), names_to = "periode", values_to = "groei") %>%
  mutate(
    variabele = "Overweight",
    jaar = as.numeric(sub(".*_", "", periode))
  )

# samenvoegen
groei_samen <- bind_rows(groei_inkomen_lang, groei_overgewicht_lang)

# grafiek met lijnen
ggplot(groei_samen, aes(x = jaar, y = groei, color = variabele)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linewidth = 0.6, linetype = "dashed") +
  scale_color_manual(values = c("Income" = "steelblue", "Overweight" = "darkorange")) +
  labs(
    title = "Growth income and overweight over time",
    x = "Year",
    y = "Growth (%)",
    color = NULL
  ) +
  theme_minimal()
```

## 3.4 Visualize spatial variation

```{r visualise_map, echo=FALSE}
#Spatial variation visualized 
combined <- merge(inkomen_landen_europa, overgewichtLanden, by = "regio")

# werken aan nieuwe grafiek
# Library
library(cartography)
library(sp)



library(cartography)
library(sp)
library(dplyr)

library(dplyr)

#data <- rbind(
 # data,
  #data.frame(regio = "Frankrijk", overgewicht2022 = NA)
#)



iso_tabel <- data.frame(
  regio = c(
    "Belgium","Bulgaria","Czechia","Denmark","Germany","Estonia",
    "Ireland","Greece","Spain","France","Croatia","Italy",
    "Cyprus","Latvia","Lithuania","Luxembourg","Hungary","Malta",
    "Netherlands","Austria","Poland","Portugal","Romania",
    "Slovenia","Slovakia","Finland","Sweden","Iceland",
    "Norway","Switzerland","Bosnia and Herzegovina",
    "North Macedonia","Serbia","Türkiye","France"
  ),
  id = c(
    "BE","BG","CZ","DK","DE","EE","IE","EL","ES","FR",
    "HR","IT","CY","LV","LT","LU","HU","MT","NL","AT",
    "PL","PT","RO","SI","SK","FI","SE","IS","NO","CH",
    "BA","MK","RS","TR","FR"
  )
)

combined <- merge(combined, iso_tabel, by = "regio")

# Cartography data laden
#data(nuts2006)



library(rnaturalearth)
library(sf)
library(dplyr)

europe <- ne_countries(scale = "medium", continent = "Europe", returnclass = "sf") |>
  select(name, iso2 = iso_a2, geometry)
europe$name[europe$iso2 == "FR"]


head(europe)
# name, iso2 (bv. "NL", "DE", "FR"), geometry (MULTIPOLYGON)

europe <- europe |>
  filter(!iso2 %in% c("RU")) 

combined$id
europe$iso2
combined <- left_join(europe, combined, join_by(iso2 == id))

combined_df <- st_drop_geometry(combined)



# Variabele numeriek maken
combined$overgewicht2022 <- as.numeric(combined$overgewicht2022)

# Kaart tekenen

combined$overgewicht2022 <- combined$overgewicht2022 %>% 
  as.numeric()





library(ggplot2)
#ggplot(combined) +
 # geom_sf(aes(fill = overgewicht2022)) +
 # theme_void()


filter(combined, combined$overgewicht2022 == ":")

combined$overgewicht2022[combined$overgewicht2022 == ":"] <- NA


library(paletteer)
paletteer_c("ggthemes::Red-Green-Gold Diverging", 30)

library(ggplot2)
library(paletteer)

kleuren <- paletteer_c("ggthemes::Red-Green-Gold Diverging", 30)

library(paletteer)

kleuren <- rev(as.character(
  paletteer_c("ggthemes::Red-Green-Gold Diverging", 30)
))

ggplot(combined) +
  geom_sf(aes(fill = overgewicht2022), color = "grey50", linewidth = 0.2) +
  scale_fill_gradientn(
    colours = kleuren,
    na.value = "lightgrey",
    name = "Overweight (%)"
  ) +
  theme_void() +
  labs(
    title = "Overweight by European country (2022)"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations, echo=FALSE, warning=FALSE}
#Sub-population variation visualized 
# 1. Pakketten laden
install.packages("viridis")   # alleen de eerste keer nodig
library(viridis)
library(tidyverse)            # bevat ggplot2, dplyr, tidyr

# 2. Inkomensgroep aanmaken in 'combined'
combined <- combined %>%
  mutate(
    inkomen2022 = as.numeric(inkomen2022),
    groep = case_when(
      inkomen2022 < 20000                         ~ "Lower",
      inkomen2022 >= 20000 & inkomen2022 < 60000  ~ "Middle",
      inkomen2022 >= 60000                        ~ "Upper"
    )
  )

# 3. Data van breed naar lang omzetten
boxplot_long <- combined %>%
  select(regio, groep, overgewicht2017, overgewicht2022, overgewicht2025) %>%
  mutate(across(starts_with("overgewicht"), as.numeric)) %>%   # <- deze regel toevoegen
  pivot_longer(
    cols = c(overgewicht2017, overgewicht2022, overgewicht2025),
    names_to = "jaar",
    values_to = "overgewicht"
  ) %>%
  mutate(
    jaar = recode(jaar,
      "overgewicht2017" = "2017",
      "overgewicht2022" = "2022",
      "overgewicht2025" = "2025"
    )
  )

# 4. De boxplot
boxplot_long %>%
  filter(!is.na(groep)) %>%
  mutate(groep = factor(groep, levels = c("Lower", "Middle", "Upper"))) %>%
  ggplot(aes(x = groep, y = overgewicht, fill = groep)) +
    geom_boxplot() +
    facet_wrap(~jaar) +
    scale_fill_viridis(discrete = TRUE, alpha = 0.6, option = "A") +
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 11)
    ) +
    ggtitle("Overweight % per income group, per year") +
    xlab("") +
    ylab("Overweight %")          # <- dit zet "Overweight %" op de y-as
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis, echo=FALSE}
#Event associated visualized 
library(tidyverse)
library(viridis)

# 1. Volledige regio-indeling — ALLE landen uit je data
noord <- c("Denmark", "Finland", "Sweden", "Estonia", "Latvia", "Lithuania",
           "Iceland", "Norway")
zuid  <- c("Greece", "Italy", "Spain", "Portugal", "Cyprus", "Malta",
           "Croatia", "Slovenia",
           "Bosnia and Herzegovina", "North Macedonia", "Albania", "Serbia")
oost  <- c("Poland", "Hungary", "Czechia", "Romania", "Bulgaria", "Slovakia")
west  <- c("Belgium", "Germany", "France", "Netherlands", "Austria",
           "Ireland", "Luxembourg", "Switzerland")

# 2. Breed -> lang, jaartal eruit, regio toekennen
inkomen_long <- combined %>%
  sf::st_drop_geometry() %>%
  mutate(across(starts_with("inkomen"), as.numeric)) %>%   # <- deze regel toevoegen
  pivot_longer(
    cols = starts_with("inkomen"),
    names_to = "jaar",
    values_to = "inkomen"
  ) %>%
  mutate(
    jaar = as.numeric(str_remove(jaar, "inkomen")),
    europa_deel = case_when(
      regio %in% noord ~ "North",
      regio %in% zuid  ~ "South",
      regio %in% oost  ~ "East",
      regio %in% west  ~ "West"
    )
  ) %>%
  filter(!is.na(europa_deel))

# 3. Mediaan inkomen per Europa-deel per jaar
#    (mediaan i.p.v. gemiddelde: minder gevoelig voor uitschieters
#     zoals Luxembourg en Switzerland)
inkomen_regio <- inkomen_long %>%
  filter(jaar >= 2006) %>%                  # 2002 weglaten: te veel gaten
  group_by(europa_deel, jaar) %>%
   summarise(med_inkomen = median(inkomen, na.rm = TRUE), .groups = "drop")

# 4. Connected scatterplot met de schuldencrisis gemarkeerd
ggplot(inkomen_regio, aes(x = jaar, y = med_inkomen,
                          color = europa_deel, group = europa_deel)) +
  annotate("rect", xmin = 2010, xmax = 2012, ymin = -Inf, ymax = Inf,
           alpha = 0.15, fill = "grey40") +
  annotate("text", x = 2011, y = Inf, label = "Schuldencrisis",
           vjust = 1.5, size = 3.5, color = "grey30") +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_viridis(discrete = TRUE, option = "D", end = 0.9) +
  scale_x_continuous(breaks = seq(2006, 2022, 4)) +
  theme_minimal() +
  labs(
    title = "Median income per European region (2006-2022)",
    subtitle = "Southern Europe in particular stagnates during the debt crisis",
    x = "Year",
    y = "Median income (\u20ac)",
    color = "Region"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
1. Zorginstituut Nederland. (z.d.). Obesitaszorg en zorg bij overgewicht. Geraadpleegd op 16 juni 2026, van Zorginstituut Nederland
2. De Witte, E. (2025, 4 september). Geld maakt niet gelukkig, maar wel gezonder volgens nieuw onderzoek. Manly. https://manly.nl/verband-inkomen-gezondheid-onderzoek
