---
title: "Hoe zwaar weegt welvaart" 
author: "Tobias Pierik: 2898032, Meike Bliek: 2892575, Vycho de Graaf: 2892702, Nynke Latham: 2862763, Anne Mak: 2898488" 
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("yaml")
#install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Title Page
<<<<<<< Updated upstream
<<<<<<< HEAD

#introduction
In this document, we will examine income and the percentage of overweight people across Europe. We are curious to see if a correlation can be found between a country's income and the percentage of overweight people. We started by adding two datasets, then cleaned and merged them into a single combined set. Following this, we created four graphs showing both the correlation between the two variables and the distribution of income and the percentage of overweight people across Europe.


<<<<<<< HEAD
=======
>>>>>>> Stashed changes
Include the tutorial group number: 1
=======
Include the tutorial group number: 1,8
>>>>>>> 261c22b67f2673b22462cb5c7f71ad44aefb7a9c

Include your tutorial lecturer's name: Fitzgerald, J.F.

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem
sociale vraag: In welke mate beinvloedt inkomen het percentage aan mensen met overgewicht in europa?

Dit is een relevant onderwerp omdat overgewicht zorgt voor gezondheidsproblemen. Door onderzoek te doen naar de correlatie tussen rijkdom en overgewicht, kunnen we beter begrijpen hoe inkomen de gezondheid in een land bepaald. Dit geeft ons een duidelijker overzicht voor de vraag of er een duidelijke geld oorzaak aanwezig is in het landelijke percentage v
-   ...

# Part 2 - Data Sourcing

## 2.1 Load in the data
Eurostat. (n.d.). Obesity rate by body mass index (BMI) (sdg_02_10) [Data set]. Geraadpleegd op 2 juni 2026, van https://ec.europa.eu/eurostat/databrowser/view/sdg_02_10/default/table?lang=en

Eurostat. (2024, 7 november). Annual full-time adjusted salary in EU grew in 2023. European Commission. https://ec.europa.eu/eurostat/en/web/products-eurostat-news/w/DDN-20241107-1




```{r}

```


```{r loading_data}
library(tidyverse)
dataset <- midwest
filter(dataset, state == "IL")
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)
De eerste dataset toont de ontwikkeling van het inkomen van de bevolking in verschillende Europese landen tussen 2002 en 2022. Over het algemeen is het inkomen in de meeste landen gestegen gedurende deze periode, hoewel de groei niet overal even sterk was. Daarnaast zijn er duidelijke verschillen zichtbaar tussen landen, wat wijst op uiteenlopende economische ontwikkelingen en welvaartsniveaus binnen Europa.
De tweede dataset geeft informatie over het percentage van de bevolking met overgewicht in verschillende Europese landen in 2017, 2020 en 2025. De gegevens laten zien dat overgewicht in veel landen een veelvoorkomend gezondheidsprobleem is. In meerdere landen is een stijgende trend zichtbaar, wat erop wijst dat het aandeel mensen met overgewicht in de loop der jaren verder toeneemt. Hierdoor blijft overgewicht een belangrijk aandachtspunt voor de volksgezondheid in Europa.

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}
#gemiddelde inkomen 
library(sf)
library(dplyr)
resultaat_inkomen <- combined %>%
   st_drop_geometry() %>%
  summarise(
    INCgem2002 = mean(suppressWarnings(as.numeric(inkomen2002, ":")), na.rm = TRUE),
    INCgem2006 = mean(suppressWarnings(as.numeric(inkomen2006, ":")), na.rm = TRUE),
    INCgem2010 = mean(suppressWarnings(as.numeric(inkomen2010, ":")), na.rm = TRUE),
    INCgem2014 = mean(suppressWarnings(as.numeric(inkomen2014, ":")), na.rm = TRUE),
    INCgem2018 = mean(suppressWarnings(as.numeric(inkomen2018, ":")), na.rm = TRUE),
    INCgem2022 = mean(suppressWarnings(as.numeric(inkomen2022, ":")), na.rm = TRUE)
  )
resultaat_inkomen
```
```{r}
groei_inkomen <- resultaat_inkomen %>%
  mutate(
    groei_2002_2006 = (INCgem2006 - INCgem2002) / INCgem2002 * 100,
    groei_2006_2010 = (INCgem2010 - INCgem2006) / INCgem2006 * 100,
    groei_2010_2014 = (INCgem2014 - INCgem2010) / INCgem2010 * 100,
    groei_2014_2018 = (INCgem2018 - INCgem2014) / INCgem2014 * 100,
    groei_2018_2022 = (INCgem2022 - INCgem2018) / INCgem2018 * 100
  ) %>%
  select(starts_with("groei_"))

groei_inkomen
```

Variable 2

```{r gen_var2}
#gemiddelde overgewicht
library(sf)
library(dplyr)
resultaat_overgewicht <- combined %>%
   st_drop_geometry() %>%
  summarise(
    OVgem2017 = mean(suppressWarnings(as.numeric(overgewicht2017, ":")), na.rm = TRUE),
    OVgem2022 = mean(suppressWarnings(as.numeric(overgewicht2022, ":")), na.rm = TRUE),
    OVgem2025 = mean(suppressWarnings(as.numeric(overgewicht2025, ":")), na.rm = TRUE)
  )
resultaat_overgewicht
```
```{r}
groei_overgewicht <- resultaat_overgewicht %>%
  mutate(
    groei_2017_2022 = (OVgem2022 - OVgem2017) / OVgem2017 * 100,
    groei_2022_2025 = (OVgem2025 - OVgem2022) / OVgem2022 * 100
  ) %>%
  select(starts_with("groei_"))

groei_overgewicht
```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}
#Spatial variation visualized 
combined <- merge(inkomen_landen_europa, overgewichtLanden, by = "regio")

# werken aan nieuwe grafiek
# Library
library(cartography)
library(sp)

# Upload data attached with the package.
#data(nuts2006)

# Now we have a geospatial object called nuts2.spdf containing the shape of european regions. We can plot it with the plot function.
# summary(nuts2.spdf)

# We also have a dataframe with information concerning every region.
# head(nuts2.df)
# Both object have a first column "id" that makes the link between them.

# Plot Europe
#plot(nuts0.spdf, border = NA, col = NA, bg = "#A6CAE0")
#plot(world.spdf, col = "#E3DEBF", border = NA, add = TRUE)
#plot(nuts0.spdf, col = "#D1914D", border = "grey80",  add = TRUE)

# Add circles proportional to the total population
# propSymbolsLayer(spdf = nuts0.spdf, df = nuts0.df,
#     var = "pop2008", symbols = "circle", col = "#920000",
#     legend.pos = "right", legend.title.txt = "Total\npopulation (2008)",
#     legend.style = "c")
# 
# # Add titles, legend...
# layoutLayer(title = "Countries Population in Europe",
#     author = "cartography", sources = "Eurostat, 2008",
#     scale = NULL, south = TRUE)
# 

library(cartography)
library(sp)
library(dplyr)

library(dplyr)

#data <- rbind(
 # data,
  #data.frame(regio = "Frankrijk", overgewicht2022 = NA)
#)

europe$name[europe$iso2 == "FR"]



iso_tabel <- data.frame(
  regio = c(
    "Belgium","Bulgaria","Czechia","Denmark","Germany","Estonia",
    "Ireland","Greece","Spain","France","Croatia","Italy",
    "Cyprus","Latvia","Lithuania","Luxembourg","Hungary","Malta",
    "Netherlands","Austria","Poland","Portugal","Romania",
    "Slovenia","Slovakia","Finland","Sweden","Iceland",
    "Norway","Switzerland","Bosnia and Herzegovina",
    "North Macedonia","Serbia","Türkiye","France"
  ),
  id = c(
    "BE","BG","CZ","DK","DE","EE","IE","EL","ES","FR",
    "HR","IT","CY","LV","LT","LU","HU","MT","NL","AT",
    "PL","PT","RO","SI","SK","FI","SE","IS","NO","CH",
    "BA","MK","RS","TR","FR"
  )
)

combined <- merge(combined, iso_tabel, by = "regio")

# Cartography data laden
#data(nuts2006)



library(rnaturalearth)
library(sf)
library(dplyr)

europe <- ne_countries(scale = "medium", continent = "Europe", returnclass = "sf") |>
  select(name, iso2 = iso_a2, geometry)

head(europe)
# name, iso2 (bv. "NL", "DE", "FR"), geometry (MULTIPOLYGON)

europe <- europe |>
  filter(!iso2 %in% c("RU")) 

combined$id
europe$iso2
combined <- left_join(europe, combined, join_by(iso2 == id))

combined_df <- st_drop_geometry(combined)



# Variabele numeriek maken
combined$overgewicht2022 <- as.numeric(combined$overgewicht2022)

# Kaart tekenen

combined$overgewicht2022 <- combined$overgewicht2022 %>% 
  as.numeric()

# hieronder komt ggplot code
#plot(nuts0.spdf,
#     border = "grey70",
#     col = "white")


# layoutLayer(
#   title = "Overgewicht per Europees land (2022)",
#   sources = "Eurostat",
#   author = "Jullie namen"
# )
# choroLayer(
#   spdf = nuts0.spdf,
#   df = combined,
#   var = "overgewicht2022",
#   col = carto.pal(pal1 = "orange.pal"),
#   border = "grey50",
#   legend.pos = "right",
#   legend.title.txt = "Overgewicht (%) 2022"
# )




library(ggplot2)
ggplot(combined) +
  geom_sf(aes(fill = overgewicht2022)) +
  theme_void()


filter(combined, combined$overgewicht2022 == ":")

combined$overgewicht2022[combined$overgewicht2022 == ":"] <- NA


library(paletteer)
paletteer_c("ggthemes::Red-Green-Gold Diverging", 30)

library(ggplot2)
library(paletteer)

kleuren <- paletteer_c("ggthemes::Red-Green-Gold Diverging", 30)

library(paletteer)

kleuren <- rev(as.character(
  paletteer_c("ggthemes::Red-Green-Gold Diverging", 30)
))

ggplot(combined) +
  geom_sf(aes(fill = overgewicht2022), color = "grey50", linewidth = 0.2) +
  scale_fill_gradientn(
    colours = kleuren,
    na.value = "lightgrey",
    name = "Overweight (%)"
  ) +
  theme_void() +
  labs(
    title = "Overweight by European country (2022)"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
#Sub-population variation visualized 
# 1. Pakketten laden
install.packages("viridis")   # alleen de eerste keer nodig
library(viridis)
library(tidyverse)            # bevat ggplot2, dplyr, tidyr

# 2. Inkomensgroep aanmaken in 'combined'
combined <- combined %>%
  mutate(
    inkomen2022 = as.numeric(inkomen2022),
    groep = case_when(
      inkomen2022 < 20000                         ~ "Arm",
      inkomen2022 >= 20000 & inkomen2022 < 60000  ~ "Midden",
      inkomen2022 >= 60000                        ~ "Rijk"
    )
  )

# 3. Data van breed naar lang omzetten
boxplot_long <- combined %>%
  select(regio, groep, overgewicht2017, overgewicht2022, overgewicht2025) %>%
  mutate(across(starts_with("overgewicht"), as.numeric)) %>%   # <- deze regel toevoegen
  pivot_longer(
    cols = c(overgewicht2017, overgewicht2022, overgewicht2025),
    names_to = "jaar",
    values_to = "overgewicht"
  ) %>%
  mutate(
    jaar = recode(jaar,
      "overgewicht2017" = "2017",
      "overgewicht2022" = "2022",
      "overgewicht2025" = "2025"
    )
  )

# 4. De boxplot
boxplot_long %>%
  filter(!is.na(groep)) %>%
  mutate(groep = factor(groep, levels = c("Arm", "Midden", "Rijk"))) %>%
  ggplot(aes(x = groep, y = overgewicht, fill = groep)) +
    geom_boxplot() +
    facet_wrap(~jaar) +
    scale_fill_viridis(discrete = TRUE, alpha = 0.6, option = "A") +
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.title = element_text(size = 11)
    ) +
    ggtitle("overgewicht % per inkomensgroep, opgesplitst per jaar") +
    xlab("") +
    ylab("overgewicht %")          # <- dit zet "BMI" op de y-as
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
#Event associated visualized 
library(tidyverse)
library(viridis)

# 1. Volledige regio-indeling — ALLE landen uit je data
noord <- c("Denmark", "Finland", "Sweden", "Estonia", "Latvia", "Lithuania",
           "Iceland", "Norway")
zuid  <- c("Greece", "Italy", "Spain", "Portugal", "Cyprus", "Malta",
           "Croatia", "Slovenia",
           "Bosnia and Herzegovina", "North Macedonia", "Albania", "Serbia")
oost  <- c("Poland", "Hungary", "Czechia", "Romania", "Bulgaria", "Slovakia")
west  <- c("Belgium", "Germany", "France", "Netherlands", "Austria",
           "Ireland", "Luxembourg", "Switzerland")

# 2. Breed -> lang, jaartal eruit, regio toekennen
inkomen_long <- combined %>%
  sf::st_drop_geometry() %>%
  mutate(across(starts_with("inkomen"), as.numeric)) %>%   # <- deze regel toevoegen
  pivot_longer(
    cols = starts_with("inkomen"),
    names_to = "jaar",
    values_to = "inkomen"
  ) %>%
  mutate(
    jaar = as.numeric(str_remove(jaar, "inkomen")),
    europa_deel = case_when(
      regio %in% noord ~ "Noord",
      regio %in% zuid  ~ "Zuid",
      regio %in% oost  ~ "Oost",
      regio %in% west  ~ "West"
    )
  ) %>%
  filter(!is.na(europa_deel))

# 3. Mediaan inkomen per Europa-deel per jaar
#    (mediaan i.p.v. gemiddelde: minder gevoelig voor uitschieters
#     zoals Luxembourg en Switzerland)
inkomen_regio <- inkomen_long %>%
  filter(jaar >= 2006) %>%                  # 2002 weglaten: te veel gaten
  group_by(europa_deel, jaar) %>%
   summarise(med_inkomen = median(inkomen, na.rm = TRUE), .groups = "drop")

# 4. Connected scatterplot met de schuldencrisis gemarkeerd
ggplot(inkomen_regio, aes(x = jaar, y = med_inkomen,
                          color = europa_deel, group = europa_deel)) +
  annotate("rect", xmin = 2010, xmax = 2012, ymin = -Inf, ymax = Inf,
           alpha = 0.15, fill = "grey40") +
  annotate("text", x = 2011, y = Inf, label = "Schuldencrisis",
           vjust = 1.5, size = 3.5, color = "grey30") +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_color_viridis(discrete = TRUE, option = "D", end = 0.9) +
  scale_x_continuous(breaks = seq(2006, 2022, 4)) +
  theme_minimal() +
  labs(
    title = "Mediaan inkomen per Europese regio (2006-2022)",
    subtitle = "Vooral Zuid-Europa stagneert tijdens de schuldencrisis",
    x = "Jaar",
    y = "Mediaan inkomen (\u20ac)",
    color = "Regio"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
